Association between trauma severity models and opportunities for improvement:

### 2022-11-02 manuscript.Rmd
fix UN cites

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r main, echo=FALSE, message = FALSE}
  source("main.R") 
```

@@

Abbreviations
=============
* ASA-PS - American Society of Anesthesiology Physical Status
* ISS - Injury Severity Score
* M&M - Morbidity and mortality
* NISS - New Injury Severity Score
* NORMIT - Norwegian Survival Prediction Model in Trauma
* OFI - Opportunities for improvement
* QI - Quality Improvement
* RTS - Revised Trauma Score
* TRISS - Trauma and Injury Severity Score
* WHO - World Health Organisation

Introduction
============
Trauma, defined as the clinical entity composed of physical injury and the body's associated response [@Gerdin2015], is one of the leading causes of death worldwide, constituting around 9% of global deaths annually [@globalburden]. The global health burden of trauma has led to it being described as a worldwide pandemic. One of the characteristics of this particular pandemic is its demographic spread; trauma is a leading cause of death for people aged 10-49 years [@globalburden] and 90% of trauma-related deaths occur in low- to middle-income countries [@whoepi]. Non-fatal outcomes also comprise a significant portion of the global health burden of disability [@globalburden], which can drastically reduce quality of life and health status in survivors of severe trauma [@globalburden; @longtermtrauma]. The UN's Sustainable Development Goals, published in 2015, included the aim of reducing the number of global deaths and injuries from road traffic accidents by half by 2020 <!-- [@spg] -->. This goal was not met, as global mortality from road traffic accidents was only reduced by 8.3% from 2010 to 2019 <!-- [@spg2020] -->. Improvements in trauma outcomes stand therefore to benefit a broader population by reducing the societal and healthcare costs of disability, not least in low- and middle-income countries.

@@

The Trauma and Injury Severity Score (TRISS), first developed in the US in 1987 [@triss] and most recently updated in 2010 [@triss2010], is the most used model worldwide [@trissbetter]. TRISS is a combination index that incorporates patient age, the Injury Severity Score (ISS), the Revised Trauma Score (RTS), and whether the injury was blunt or penetrating [@triss]. The resulting value is a percentage that estimates the probability of survival (Ps). The ISS gives a score based on the three anatomical regions with the most severe injury. A limitation of the ISS is that it does not account for situations where one anatomical region has multiple injuries [@trissbetter]. The RTS is based on the physiological parameters GCS, systolic blood pressure, and respiratory rate. The GCS cannot be estimated in situations where the patient has already been intubated, since a valid verbal score cannot be ascertained, which gives TRISS limited value in estimating Ps in certain populations [@trissbetter].

@@

KUH is located in Solna, Sweden. Karolinska University Hospital's trauma centre is equivalent to a level 1 trauma centre according to American College of Surgeons standards [@nksstats]. Karolinska University Hospital is the primary trauma centre for the region of Stockholm and also has agreements with several other regions to provide trauma care for patients with severe trauma. Karolinska University Hospital is thereby the major trauma centre for a population of almost 3 million people [@nksstats].

@@

``` {r, echo = FALSE,} 
  ofi_diagram() ## Only works in HTML
```

@@

```{r, echo = FALSE}
  exclusion_flowchart(counts) ## Only works in HTML
```

@@

```{r, echo = FALSE} 
  table.one ## will be improved later
```

@@

```{r, echo = FALSE}
plot(t.stats[['roc']], main="Fig 3. TRISS Receiver Operating Characteristic Curve.", colorize = TRUE)
```

```{r, echo = FALSE}
t.plot.acc <- plot(t.stats[['acc']], main="Fig 4. TRISS Accuracy Curve.")
```

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

```{r main, echo=FALSE, message = FALSE}
  source("main.R") 
```

@@

Abbreviations
=============
* ASA-PS - American Society of Anesthesiology Physical Status
* ISS - Injury Severity Score
* M&M - Morbidity and mortality
* NISS - New Injury Severity Score
* NORMIT - Norwegian Survival Prediction Model in Trauma
* OFI - Opportunities for improvement
* PS - Probability of Survival score
* QI - Quality Improvement
* RTS - Revised Trauma Score
* TRISS - Trauma and Injury Severity Score
* WHO - World Health Organisation

Introduction
============
Trauma, defined as the clinical entity composed of physical injury and the body's associated response [@Gerdin2015], is one of the leading causes of death worldwide, constituting around 9% of global deaths annually [@globalburden]. The global health burden of trauma has led to it being described as a worldwide pandemic. One of the characteristics of this particular pandemic is its demographic spread; trauma is a leading cause of death for people aged 10-49 years [@globalburden] and 90% of trauma-related deaths occur in low- to middle-income countries [@whoepi]. Non-fatal outcomes also comprise a significant portion of the global health burden of disability [@globalburden], which can drastically reduce quality of life and health status in survivors of severe trauma [@globalburden; @longtermtrauma]. The UN's Sustainable Development Goals, published in 2015, included the aim of reducing the number of global deaths and injuries from road traffic accidents by half by 2020 [@sdg]. This goal was not met, as global mortality from road traffic accidents was only reduced by 8.3% from 2010 to 2019 [@sdg20]. Improvements in trauma outcomes stand therefore to benefit a broader population by reducing the societal and healthcare costs of disability, not least in low- and middle-income countries.

@@

The Trauma and Injury Severity Score (TRISS), first developed in the US in 1987 [@triss] and most recently updated in 2010 [@triss2010], is the most used model worldwide [@trissbetter]. TRISS is a combination index that incorporates patient age, the Injury Severity Score (ISS), the Revised Trauma Score (RTS), and whether the injury was blunt or penetrating [@triss]. The resulting value is a percentage that estimates the probability of survival (Ps). The ISS gives a score based on the three anatomical regions with the most severe injury. A limitation of the ISS is that it does not account for situations where one anatomical region has multiple injuries [@trissbetter]. The RTS is based on the physiological parameters GCS, systolic blood pressure, and respiratory rate. The GCS cannot be estimated in situations where the patient has already been intubated, since a valid verbal score cannot be ascertained, which gives TRISS limited value in estimating probability of survival in certain populations [@trissbetter].

@@

Karolinska University Hospital is located in Solna, Sweden. Karolinska University Hospital's trauma centre is equivalent to a level 1 trauma centre according to American College of Surgeons standards [@nksstats]. Karolinska University Hospital is the primary trauma centre for the region of Stockholm and also has agreements with several other regions to provide trauma care for patients with severe trauma. Karolinska University Hospital is thereby the major trauma centre for a population of almost 3 million people [@nksstats].

@@

``` {r} 
  ofi_diagram() ## Only works in HTML
```

@@

```{r}
  exclusion_flowchart(counts) ## Only works in HTML
```

@@

```{r} 
  table.one ## will be improved later
```

@@

```{r}
plot(t.stats[['roc']], main="Fig 3. TRISS Receiver Operating Characteristic Curve.", colorize = TRUE)
```

```{r}
t.plot.acc <- plot(t.stats[['acc']], main="Fig 4. TRISS Accuracy Curve.")
```
